---
title: "Methodological Note — Alert Trends and Adequacy Analysis, EVD17 Response"
subtitle: "Describing the implemented adequacy metrics and trend tests"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    number-sections: true
    theme: flatly
  pdf:
    toc: true
    number-sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
```

## Purpose

This note documents the trend-analysis workflow implemented in `R/02_alert_trends.R`.
It explains how the script turns the threshold synthesis from script 01 into weekly adequacy measures, smoothed counts, recent rankings, and Mann-Kendall trend tests.

The operational question is:

> Are health zones generating the expected volume of validated alerts, and is that volume changing over time?

---

## Data inputs

| Input | Source | Role |
|---|---|---|
| Cleaned EVD line list | Latest file under `../DataCleaning/data/Output/` | Validated alerts from 2026 onward |
| Threshold synthesis table | `output/01_thresholds_synthesis.rds` | HZ- and window-specific threshold values |

The script loads both inputs, keeps validated alerts only, converts notification dates to `Date`, and restricts the analysis to 2026 onward.

---

## Alert counting and threshold joining

Validated alerts are joined to the threshold synthesis and summarised by time window (week start date) and health zone.
The analysis keeps separate counts for case alerts and death alerts, and it derives a `total_alerts` series for trend testing.

The threshold columns are normalized to consistent names before adequacy is computed.

---

## Alert Adequacy Index (AAI)

For each HZ and week, the script computes midpoint thresholds:

$$
T_{case}^{mid} = \frac{T_{case}^{lower} + T_{case}^{upper}}{2}
$$

$$
T_{death}^{mid} = \frac{T_{death}^{lower} + T_{death}^{upper}}{2}
$$

Adequacy ratios are then defined as:

$$
A_{case} = \frac{case\_alerts}{T_{case}^{mid}}, \qquad
A_{death} = \frac{death\_alerts}{T_{death}^{mid}}
$$

The Alert Adequacy Index is the mean of the available adequacy ratios:

$$
AAI = \text{mean}(A_{case}, A_{death})
$$

When no valid threshold midpoint is available, AAI is set to missing.

### Interpretation

- `AAI = 1` indicates alert volume matches the threshold midpoint.
- `AAI < 1` indicates fewer alerts than expected.
- `AAI > 1` indicates more alerts than expected.

The script classifies adequacy using the following cut points:

| Category | Condition |
|---|---|
| Under-alerting | `AAI < 0.75` |
| Adequate | `0.75 <= AAI <= 2.0` |
| Over-alerting | `AAI > 2.0` |

For the recent summary table, the script averages AAI over the most recent 3 weeks.

---

## Temporal smoothing

The trend script applies a right-aligned 3-week rolling mean to case and death alerts within each HZ.
This smooths short-term noise while retaining the current week in the rolling window.

The smoothed counts are stored as `case_alerts_3w` and `death_alerts_3w`.

---

## Trend testing

The script applies the Mann-Kendall test to `total_alerts` within each HZ.
The test is run only when there are at least 4 weeks with AAI data in the zone.

For each HZ, the script extracts:

- Kendall’s tau
- the p-value
- a simple trend label:
  - Increasing: `p < 0.05` and `tau > 0`
  - Decreasing: `p < 0.05` and `tau < 0`
  - Stable: all other non-missing cases
  - Unknown: failed test or missing result

The same procedure is repeated for the most recent 6 weeks to capture short-term movement.

---

## Recent adequacy ranking

For the most recent 3 weeks, the script computes mean AAI by HZ, assigns an adequacy category, joins the long-run trend labels, and sorts the result by mean AAI.
This produces a ranking table that highlights HZs that are currently under-alerting or over-alerting.

---

## Outputs

The script saves:

- `02_trends_smooth.rds`
- `02_trends_smooth_adeq.rds`
- `02_recent_adequacy.xlsx`
- `trends_smooth.xlsx`
- `trends_smooth_adeq.xlsx`
- `02_alert_trends.pdf`

The PDF includes an overall alert trend plot and a faceted view for the highest-priority under-alerting HZs.

---

## Glossary

The column-level glossary for `02_trends_smooth_adeq.rds` is included below.

```{r include_trends_glossary, echo=FALSE, results='asis'}
cat(readLines("../glossaries/02_trends_glossary_en.md"), sep = "\n")
```

---

## Caveats and limitations

- The AAI depends on the quality of the threshold synthesis from script 01.
- Right-aligned rolling means introduce lag.
- Mann-Kendall detects monotonic trends, not abrupt changes or seasonality.
- The current adequacy thresholds are operational cut points, not formal clinical standards.
- HZs with sparse data can have unstable trend labels.
