---
title: "Methodological Note — Weekly Alert Thresholds, EVD17 Response"
subtitle: "Documenting the implemented threshold synthesis and its assumptions"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    number-sections: true
    theme: flatly
  pdf:
    toc: true
    number-sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
```

## Purpose

This note documents the threshold-generation workflow implemented in `R/01_alert_thresholds.R`.
It describes how the script derives weekly alert thresholds per health zone (HZ) from three sources: baseline mortality, the historical Beni alert dataset, and case-derived expectations from the current line list.

The operational question is:

> Given the population size and current epidemiology of a health zone, how many validated alerts for cases and deaths in a week are unusual enough to warrant review?

The script combines multiple imperfect sources because no single method is reliable in this setting. The goal is to produce a defensible threshold table and preserve the intermediate estimates used to build it.

---

## Data inputs

| Input | Source | Role |
|---|---|---|
| Cleaned EVD line list | Latest file under `../DataCleaning/data/Output/` | Current cases, deaths, CFR, detection, Rt, and recent alert counts |
| Beni historical alerts | `data/Alerts/Alert_Beni.rds` | Historical benchmark during low-positivity periods |
| Population by HZ | `../data/PopulationParAge/DRC_ZS_DHIS2_Pop_2024.xlsx` | Denominator for per-capita thresholds |
| Crude mortality rate | 8.3 per 1,000 per year | Baseline mortality assumption used for the first approach |

Population is projected from the 2024 DHIS2 baseline to the current year using compound annual growth in the script.

---

## Approach A: Baseline mortality

This approach converts crude mortality into an expected weekly death count per HZ.

For a health zone with population $P$ and crude mortality rate $m$ per 1,000 per year, the script computes:

$$
E[D_w] = \frac{P \times m}{1000} \times \frac{7}{365}
$$

The lower and upper thresholds are set as fixed 10% bounds around this expected weekly value in the implemented code:

$$
T_A^{lower} = 0.9 \times E[D_w], \qquad T_A^{upper} = 1.1 \times E[D_w]
$$

These are death-only thresholds.

---

## Approach B: Beni historical benchmark

This approach uses validated alerts in the historical Beni dataset as an empirical reference.

The implemented workflow is:

1. Restrict to validated alerts.
2. Aggregate by HZ and week.
3. Compute positivity as:

$$
\text{positivity} = \frac{\text{positive samples}}{\text{sampled validated alerts}}
$$

For Beni, sampled alerts are proxied by non-missing laboratory results.
4. Define optimal weeks as those with positivity below 10%.
5. Within optimal weeks, compute weekly alert rates per 100,000 population.
6. Retain HZs with at least 50 validated alerts and case-alert rates in the top quartile.
7. Use the Beni reference values from the best-performing HZs as the benchmark rates.
8. Scale those rates to each target HZ population.

The benchmark values are stored as case and death rates, then converted to HZ-specific alert counts by multiplying by population / 100,000.

---

## Approach C: Case-derived expectations

This approach uses the current line list to estimate the number of alerts that should be expected from the outbreak context in each HZ. It is the only approach that adapts to the *current epidemic state* of each zone.

Approach C chains five building blocks, documented below:

1. **Epidemiological parameters** estimated per HZ (CFR, back-calculated CFR, growth rate, $R_t$, SAR).
2. **Detection rate from epi-links** — the proportion of confirmed cases found through active contact tracing.
3. **Detection rate from back-calculation** — the proportion of true cases actually detected, inferred from observed deaths.
4. **Combination** of the two detection estimates into a single rate used to back-calculate true case counts.
5. **Poisson offset multiplier model** that translates estimated true cases into expected validated-alert counts, producing $\beta_c$ and $\beta_d$.

The case-derived thresholds are therefore window-specific and HZ-specific, and they are merged with the baseline and Beni-derived thresholds in the final synthesis table.

### C.1 Epidemiological parameters

For each HZ, the script estimates:

| Parameter | Function | Definition |
|---|---|---|
| CFR (`cfr_used`) | `compute_cfr_by_hz()` | Deaths among confirmed cases / confirmed cases; pooled CFR used when HZ has < 5 confirmed cases |
| Back-calculated CFR | `compute_cfr_backcalc_by_hz()` | See C.3 |
| Growth rate $r$ | `compute_growth_rate_by_hz()` | Exponential growth rate from `incidence::fit()` in rolling 7-day windows; pooled across all HZs as fallback |
| $R_t$ | `compute_rt_hz()` | `EpiEstim` parametric SI (mean 12, sd 5 days) over last 21 days of onset dates; pooled $R_t$ as fallback for sparse HZs |
| Contacts per case ($k$) | `compute_contacts_per_case_by_hz()` | Followed contacts ÷ confirmed cases within the analysis window; HZ-specific when the HZ has ≥ 5 confirmed cases and a positive ratio, otherwise the overall pooled ratio |
| SAR | Derived | $R_t / k$ using the estimated contacts per case; coerced to 0.1 when missing |
| Expected deaths | Derived | $\hat{C}_{\text{true}} \times \text{CFR}$ |
| Expected secondary cases | Derived | $\hat{C}_{\text{true}} \times k \times \text{SAR}$ (equals $\hat{C}_{\text{true}} \times R_t$ since SAR $= R_t / k$) |

### C.2 Epi-link detection method

The epi-link method estimates the detection rate from the proportion of confirmed cases that were found through active contact tracing (marked by an epidemiologic link), among confirmed cases with valid epidemiologic-link data.

$$
\rho_{\text{epilink}} = \frac{n_{\text{epilink}}}{n_{\text{conf, valid epilink}}}
$$

where $n_{\text{epilink}}$ is the number of confirmed/positive cases with `alert_lien_epidemiologic == "Oui"`, and $n_{\text{conf, valid epilink}}$ is the number of confirmed cases with non-missing epidemiologic-link data.

When $n_{\text{conf, valid epilink}} < 5$, the HZ-specific rate is replaced by the pooled (all-HZ) detection rate:

$$
\rho_{\text{epilink, pooled}} = \frac{\sum_{\text{all HZ}} n_{\text{epilink}}}{\sum_{\text{all HZ}} n_{\text{conf, valid epilink}}}
$$

The under-detection rate is $1 - \rho_{\text{epilink}}$.

**Rationale.** A high proportion of epi-linked cases suggests active case-finding is working well; a low proportion suggests many cases are arriving through passive routes and may reflect under-detection.

### C.3 Back-CFR detection method

The back-CFR method estimates the detection rate from observed deaths, in two stages.

**Stage 1 — Back-calculated CFR.** The back-calculated CFR enlarges the denominator beyond confirmed cases alone by including eligible deaths that were never laboratory-confirmed:

$$
\text{CFR}_{\text{backcalc}} = \frac{D_{\text{eligible}}}{N_{\text{confirmed}} + D_{\text{eligible, unconfirmed}}}
$$

where $D_{\text{eligible}}$ is the number of *eligible deaths* — observed deaths (alert status or final status = "Décédé") with a positive or unavailable laboratory result — and $D_{\text{eligible, unconfirmed}}$ is the subset of those deaths that are not also confirmed cases. Exact binomial 95% confidence intervals are computed. HZs with denominator < 5 fall back to the pooled CFR.

**Stage 2 — Detection rate from deaths.** The detection rate is the ratio of detected confirmed cases to the estimated true case count, where true cases are back-calculated from observed deaths:

$$
\rho_{\text{backcalc}} = \frac{N_{\text{detected}}}{\hat{C}_{\text{true}}}
$$

with the true case estimate:

$$
\hat{C}_{\text{true}} = \frac{D_{\text{observed}} \times (1 + r/\beta)^{\alpha}}{\text{CFR}_{\text{backcalc}}}
$$

**Delay-distribution adjustment.** The term $(1 + r/\beta)^{\alpha}$ corrects the right-censoring bias that arises because deaths lag behind case onsets during a growing or declining epidemic. The onset-to-death delay is modelled as Gamma-distributed with:

$$
\alpha = \left(\frac{\mu}{\sigma}\right)^2, \qquad \beta = \frac{\mu}{\sigma^2}
$$

where $\mu = 11.37$ days and $\sigma = 5.41$ days (literature values for EVD onset-to-death delay). The exponential growth rate $r$ is estimated per HZ per rolling window. When $r > 0$ (growing epidemic), the adjustment factor exceeds 1, inflating the estimated true case count to account for deaths that have not yet been observed; when $r < 0$, it is less than 1.

Confidence intervals on $\rho_{\text{backcalc}}$ combine the CFR confidence bounds with the growth-rate confidence bounds to produce `detection_rate_low` (pessimistic) and `detection_rate_high` (optimistic).

### C.4 Combining detection estimates

The two detection estimates are combined into a single rate per HZ:

1. **Validity filter.** Only detection rates satisfying $0 < \rho \leq 1$ are retained. Rates outside this range (e.g., back-calc estimates > 1 from sparse data) are set to missing.

2. **Averaging.** The combined detection rate is the arithmetic mean of valid sources:

$$
\rho_{\text{combined}} = \frac{1}{S} \sum_{s \in \{\text{backcalc},\, \text{epilink}\}} \rho_s
$$

where $S$ is the number of valid sources (0, 1, or 2). If neither source is valid, the combined rate is missing and downstream true-case estimates are missing.

3. **True case estimation.** Estimated true cases in the recent window:

$$
\hat{C}_{\text{true, recent}} = \frac{C_{\text{detected, recent}}}{\rho_{\text{combined}}}
$$

where $C_{\text{detected, recent}}$ is the count of confirmed/positive cases in the most recent 7-day window. For the most recent window, this count may be replaced by a growth-model projection (`predicted_last_count`) to mitigate reporting-delay right-censoring.

4. **Source tracking.** The output records which methods contributed (`detection_rate_source_names`: "backcalc", "epilink", or "backcalc,epilink") and the combination status (`detection_rate_combined_status`: "averaged", "single_source", or "no_valid_detection_rate").

### C.5 Rolling-window architecture

Approach C produces time-varying thresholds using non-overlapping windows:

- **Window generation.** Non-overlapping 7-day windows are generated backward from the most recent confirmed-case onset date. Each window is identified by a `threshold_time_key` (its start date) and carries a `recent_case_window_start` / `recent_case_window_end` pair (the exposure period) and a `threshold_valid_from` / `threshold_valid_to` pair (the period during which the threshold applies).
- **Single time axis (harmonized, 2026-08-16).** This analysis grid is the *only* weekly time axis used in the alert pipeline. Alert counts, confirmed-case exposure, nowcast deltas, and the Poisson multiplier model are all binned onto these grid windows (`bin_dates_by_grid()` in `alert_helpers/window_utils.R`); Monday-calendar-week binning is not used for live indicators. The grid is data-anchored (windows end on the latest confirmed-case date), so its start weekday follows the data — Sunday in the current snapshot — rather than a fixed calendar convention.
- **Reporting-delay guard.** The last `n_copy_last = 1` window copies its growth-rate estimate (and consequently its multiplier estimates) from the preceding window, because the most recent days are subject to right-censoring from notification delays.
- **Pooled fallback.** A pooled (all-HZ combined) growth rate is computed for each window date and used when the HZ-specific estimate is unavailable (too few cases, fit failure).
- **Analysis bounds.** Windows are bounded by `analysis_start_date` (max of earliest confirmed-case date, 2026-04-01) and `analysis_end_date` (min of latest confirmed-case date, today).

### C.6 Poisson offset multiplier model

For each HZ and threshold window, a Poisson GLM with log-link is fitted to estimate the alert multipliers $\beta_c$ (case-alert) and $\beta_d$ (death-alert):

$$
\log E[A_w] = \log(E_w) + \beta_0
$$

where $A_w$ is the validated-alert count in grid window $w$ (alive alerts for $\beta_c$, dead alerts for $\beta_d$), and $E_w$ is the expected exposure. The multiplier is $\beta = \exp(\beta_0)$.

**Window binning.** Validated alerts, confirmed-case exposure, and nowcast deltas are binned onto the same analysis grid windows used for threshold reporting (see C.5) — not Monday calendar weeks — so each grid window contributes exactly one observation to the model and the estimated rate applies to the same 7-day periods as the thresholds it feeds. For a given threshold window, the model is fitted on the cumulative grid series observed up to that window end (one observation per grid window, no lookahead).

**Exposure allocation.** The HZ-level total estimated true cases is allocated across grid windows in proportion to detected confirmed cases:

$$
E_w^{\text{case}} = \hat{C}_{\text{true, recent}} \times \frac{c_w}{\sum_{w} c_w}, \qquad E_w^{\text{death}} = \hat{C}_{\text{true, recent}} \times \text{CFR} \times \frac{c_w}{\sum_{w} c_w}
$$

where $c_w$ is the detected confirmed-case count in grid window $w$.

**Dispersion handling.** The dispersion estimate is $\phi = \text{deviance} / \text{df}_{\text{residual}}$ from the Poisson fit. When $\phi$ exceeds the overdispersion threshold (default $\phi > 1.5$), the model is refitted as a negative-binomial GLM with the same log-link and offset, and $\beta$ with its confidence interval are taken from that fit (the additional dispersion parameter $\theta$ is estimated from the data). For $\phi \le 1.5$ — or when the negative-binomial fit does not converge — standard errors are scaled by $\sqrt{\max(1, \phi)}$, producing quasi-Poisson-style intervals.

**Fallback.** If fewer than `min_model_weeks = 2` grid windows have positive exposure, the GLM is not fitted. A simple rate ratio $\beta = \sum A_w / \sum E_w$ is used instead, with exact Poisson confidence intervals. If total exposure is zero or non-finite, the multiplier is missing and a pooled multiplier (computed across all HZs from validated alerts over estimated true cases) is substituted at synthesis time.

**Interpretation.** $\beta > 1$ means the alert system generates more validated alerts than predicted by the estimated true burden (possibly including non-EVD alerts); $\beta < 1$ means under-ascertainment relative to estimated true cases.

**Threshold construction.** Case-derived thresholds are purely data-driven from estimated case burden and Poisson offset multipliers (without adding historical Approach B baseline floors, which are integrated separately in the threshold synthesis phase):

$$
T_C^{\text{point}} = \beta \times E, \qquad T_C^{\text{lower}} = \beta_{\text{low}} \times E, \qquad T_C^{\text{upper}} = \beta_{\text{high}} \times E
$$

where $E$ is the expected count (`estimated_true_cases_recent` for cases, `expected_deaths` for deaths), and $\beta, \beta_{\text{low}}, \beta_{\text{high}}$ are the central estimate and uncertainty bounds from the offset GLM (Poisson, or negative-binomial when the fit is overdispersed).

---

## Threshold synthesis

The final synthesis table combines the three approaches into a single set of thresholds per HZ and threshold window.

In the implemented code:

- `Alert_case_lower` and `Alert_case_upper` average the Beni benchmark with the case-derived lower and upper thresholds.
- `Alert_death_lower` and `Alert_death_upper` average the baseline mortality threshold, the Beni benchmark, and the case-derived threshold.
- Central thresholds are then computed as the midpoint between the lower and upper bounds.
- All numeric threshold columns are rounded to whole numbers before saving.

The script also saves a list of intermediate parameter tables, including CFR, detection, growth, recent cases, multipliers, Rt/SAR, baseline mortality, Beni benchmark summaries, and case-derived threshold components.

---

## Outputs

The script writes the following files to the dated output directory:

- `01_thresholds_synthesis.rds`
- `01_thresholds_synthesis.xlsx`
- `synthesis_short.xlsx`
- `01_intermediate_parameters.rds`
- `EpiSize_BDV.xlsx`

These outputs are the inputs to the trend analysis script.

---

## Glossary

The column-level glossary for `01_thresholds_synthesis.rds` is included below.

```{r include_thresholds_glossary, echo=FALSE, results='asis'}
cat(readLines("../glossaries/01_thresholds_glossary_en.md"), sep = "\n")
```

---

## Caveats and limitations

- Population figures are projected from 2024 and may not reflect displacement or local demographic change.
- The crude mortality rate is applied as a fallback assumption and is not zone-specific.
- `k`, the contacts per case, is estimated from the cleaned contact follow-up line list and the EVD line list over the analysis window (HZ-specific where reliable, pooled otherwise). The estimate is sensitive to the completeness of contact registration; HZs with incomplete contact follow-up registration fall back to the pooled ratio.
- The Beni benchmark depends on the definition of low-positivity periods and the quality of the historical dataset.
- The case-derived thresholds depend on sparse-data estimates and rolling-window model fitting, so uncertainty can be substantial in some HZs.
- The epi-link method assumes that `alert_lien_epidemiologic == "Oui"` indicates active case-finding; misclassification of this field biases the detection estimate.
- The back-CFR method assumes the onset-to-death delay distribution (Gamma, mean 11.37, sd 5.41 days) is correct for this setting; a different delay would change the delay-adjustment factor.
- The two detection rates are combined by simple unweighted averaging; no precision-weighting or hierarchical model is applied, so a noisy single-source estimate can outweigh a more stable paired estimate.
- The Poisson offset model assumes weekly true cases can be approximated by proportionally allocating the window total across epiweeks based on detected confirmed-case counts; if detection varied substantially across weeks within a window, this allocation is biased.
- The script is documentation of the implemented workflow, not a claim that each component is equally precise.
